---
title: "Getting Started with biomarkerPanels"
author: "Package Authors"
output:
  BiocStyle::html_document:
    toc: true
vignette: >
  %\VignetteIndexEntry{Getting Started with biomarkerPanels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
if (!requireNamespace("biomarkerPanels", quietly = TRUE)) {
  if (requireNamespace("pkgload", quietly = TRUE)) {
    pkgload::load_all("..", quiet = TRUE)
  } else {
    stop(
      "`biomarkerPanels` is not installed. Install the package or run this vignette ",
      "from within the source tree with pkgload available.",
      call. = FALSE
    )
  }
}
library(biomarkerPanels)
```

## Introduction

`biomarkerPanels` helps researchers build compact biomarker signatures when
multiple cohorts are available and the number of measured features greatly
exceeds the number of samples. The package delegates the heavy lift of
multi-objective search to `rmoo` (NSGA-II/NSGA-III) while exposing a friendly
interface for selecting metrics, restricting the feature pool, and
summarising the resulting Pareto front.

This vignette walks through three steps:

1. Prepare multi-cohort expression data.
2. Assemble objectives from the metric registry.
3. Run `optimize_panel()` to discover candidate panels and inspect the results.

## 1. Load a Multi-cohort Dataset

For demonstration we ship a small synthetic dataset in the test fixtures. In a
package build, these files would live in `inst/extdata/`; here we reuse the
fixture directly.

```{r load-fixture}
fixture_candidates <- c(
  system.file("extdata", "fake_gene_expression.Rds", package = "biomarkerPanels"),
  system.file("test-data", "fake_gene_expression.Rds", package = "biomarkerPanels"),
  file.path("..", "tests", "data", "fake_gene_expression.Rds"),
  file.path("..", "..", "tests", "data", "fake_gene_expression.Rds")
)
fixture_candidates <- fixture_candidates[nzchar(fixture_candidates)]
fixture_path <- NULL
for (candidate in fixture_candidates) {
  if (file.exists(candidate)) {
    fixture_path <- candidate
    break
  }
}
if (is.null(fixture_path)) {
  stop(
    "Unable to locate `fake_gene_expression.Rds`. ",
    "Regenerate it via `data-raw/simulate_gene_expression.R` if needed.",
    call. = FALSE
  )
}
cohorts <- readRDS(fixture_path)

lapply(cohorts$x_list, dim)
colnames(cohorts$x_list[[1]])[1:6]
```

Each element of `x_list` is a log-scale expression matrix with the same feature
ordering across cohorts. The matching `y_list` contains binary outcome labels
stored as factors with levels `"No"` and `"Yes"`.

## 2. Configure Objectives

Metrics are registered in an extensible registry. You can inspect the
available metrics with `metric_registry()`.

```{r registry}
names(metric_registry())
```

To emphasise clinical utility we combine sensitivity, specificity, and a panel
size penalty. We also tighten the classification probability cutoff for sensitivity.

```{r objectives}
objectives <- define_objectives(
  metrics = c("sensitivity", "specificity", "num_features"),
  params = list(sensitivity = list(cutoff_prob = 0.6))
)
objectives[["sensitivity"]]$direction
```

When multiple cohorts are available you can layer on transfer-oriented
objectives that operate on cohort-level sensitivities, calibration, and feature
shifts. These metrics automatically receive cohort labels during optimisation.

```{r transfer-objectives, eval=FALSE}
transfer_objectives <- define_objectives(
  metrics = c(
    "min_cohort_auc",
    "cohort_auc_gap",
    "cohort_auc_var",
    "max_cohort_brier"
  )
)
```

### A Rule-Out Recipe

Rule-out diagnostics pin the operating point at high sensitivity and then push
specificity as far as it will go. There is no dedicated helper for this — it is
an ordinary `define_objectives()` call in which
`specificity_at_sensitivity`'s `target_sensitivity` and `pauc`'s `sens_floor`
both name the same operating point.

```{r ruleout-objectives, eval=FALSE}
ruleout_objectives <- define_objectives(
  metrics = c("specificity_at_sensitivity", "pauc", "min_cohort_auc",
              "num_features"),
  params = list(
    specificity_at_sensitivity = list(target_sensitivity = 0.90),
    pauc = list(sens_floor = 0.90, partial_auc_correct = TRUE)
  )
)
```

No hard constraint is needed: because both metrics are already evaluated at the
0.90 sensitivity floor, candidates that cannot reach it score poorly on their
own. Add a `min_metric_constraint("sensitivity", threshold = 0.9)` only if you
want infeasible candidates discarded outright rather than ranked last.

### Enforce Minimum Standards

Some clinical settings demand a hard floor on performance before optimising
another objective. Supply constraint descriptors—such as
`min_metric_constraint()`—via the `constraints` argument to
`optimize_panel()`. The optimiser will discard any candidate failing the
constraint.

```{r constraints, eval=FALSE}
constraints <- list(
  min_metric_constraint("sensitivity", threshold = 0.9,
                        params = list(cutoff_prob = 0.6))
)
```

## 3. Run NSGA-II Optimisation

`optimize_panel()` accepts single matrices or lists of cohorts. We restrict the
feature pool to keep the search lightweight for this vignette.

```{r optimise, eval=FALSE}
set.seed(9001)
result <- optimize_panel(
  x = cohorts$x_list,
  y = cohorts$y_list,
  objectives = objectives,
  max_features = 5,
  feature_pool = colnames(cohorts$x_list[[1]])[1:50],
  nsga_control = list(popsize = 64, generations = 60)
)
```

The result is a `BiomarkerPanelResult` object that stores the selected features,
objective scores, and metadata about the optimisation run.

```{r result-structure, eval=FALSE}
slotNames(result)
result@features
result@metrics
head(result@objectives)
```

### Held-Out Partitioning

The multi-cohort call above scores every candidate in-sample by default. Pass
`fitness_mode` together with `train_ratio` (and `val_ratio`) to carve off a
stratified partition per cohort and score candidates against it instead:
`fitness_mode = "loco"` scores each candidate on left-out cohorts,
`"within_cohort_val"` and `"within_cohort_rotating"` score on a held-back
validation share, and `"cv"` (the overall default) k-fold cross-validates
within the training partition. Whenever `train_ratio + val_ratio < 1`, the
remaining share is stored on `result@control$heldout_x` / `heldout_y` /
`heldout_cohort` for `calibrate_panel()` to use afterwards.

### Visualise Trade-offs

We can visualise Pareto-efficient solutions with `plot_pareto_front()`. Pick any
pair of objectives and extract the corresponding columns.

```{r pareto, eval=FALSE}
library(ggplot2)
pareto_df <- subset(result@objectives, objective %in% c("sensitivity", "specificity"))
pareto_df <- reshape(
  pareto_df,
  idvar = "solution_id",
  timevar = "objective",
  direction = "wide"
)
plot_pareto_front(
  data.frame(
    objective_x = pareto_df$value.sensitivity,
    objective_y = pareto_df$value.specificity,
    label = paste0("Solution ", pareto_df$solution_id)
  ),
  xlab = "Sensitivity",
  ylab = "Specificity"
)
```

## Next Steps

- Explore additional metrics by registering custom ones with
  `register_metric()`.
- Run `BiocCheck::BiocCheck()` and `R CMD check` before submitting to
  Bioconductor.
